Empirical Analysis of Performance Bottlenecks in Graph Neural Network Training with GPUs

### 2020-12-26 NEUROC/sec2-review-of-graph-neural-network.tex
修改了sec2

\begin{table}[h]
    \caption{Notations}
    \label{tab:notations}
    \centering
    \small
    \begin{tabular}{p{3em}lp{35em}}
        \toprule
       Category & Symbol & Meaning \\
        \midrule
        \multirow[c]{4}{3em}{Graph Structure}& $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ & The simple undirected input graph with the vertex set $\mathcal{V}$ and the edge set $\mathcal{E}$. \\
        & $v_x$ & The $x$-th vertex of the input graph. \\
        & $e_{y,x}$ & The edge pointing from $v_y$ to $v_x$ of the input graph. \\
        & $\mathcal{N}(v_x)$ & The adjacency set of $v_x$ in the input graph. \\ 
        & $\bar{d}$ & The average degree of the input graph. \\ \midrule
        \multirow[c]{6}{3em}{GNN Definition}& $L$ & The number of GNN layers. \\
        & $K$ & The number of heads in a GNN layer. \\
        & $\phi^l$ & The messaging function of the GNN layer $l$. \\
        & $\Sigma^l$ & The aggregation function of the GNN layer $l$. \\
        & $\gamma^l$ & The vertex updating function of the GNN layer $l$. \\ 
        & $\phi^{l,i}$ / $\Sigma^{l,i}$ / $\gamma^{l,i}$ & The messaging/aggregation/updating function of the $i$-th sub-layer of the GNN layer  $l$.\\
        & $\textcolor{blue}{\boldsymbol{W}^l, \boldsymbol{W}^{(k)}/\boldsymbol{b}, \boldsymbol{a}}$ & The matrices/vectors represented by the blue characters are the weight matrices/vectors that need to be learned in the GNN. \\  \midrule
        \multirow[c]{8}{3em}{Vector}& $\boldsymbol{v}_x$ & The feature vector of the vertex $v_x$. \\
        & $\boldsymbol{e}_{y,x}$ & The feature vector of the edge $e_{y,x}$.  \\
        & $\boldsymbol{h}_x^{l}$ &  The {input} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$. \\
        & $\boldsymbol{h}_x^{l+1}$ &  The {output} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$.\\
        & $\boldsymbol{m}_{y,x}^l$ & The message vector of the edge $e_{y,x}$ outputted by $\phi^l$ of the GNN layer $l$. \\
        & $\boldsymbol{s}_{x}^l$ & The aggregated vector of the vertex $v_x$ outputted by $\Sigma^l$ of the GNN layer $l$. \\
        & $\boldsymbol{h}_{x}^{l,i}$ / $\boldsymbol{m}_{y,x}^{l,i}$ / $\boldsymbol{s}_{x}^{l,i}$ & The hidden/message/aggregated vector of the vertex $v_x$ outputted by $\gamma^{l,i}$/$\phi^{l,i}$/$\Sigma^{l,i}$ of the $i$-th sub-layer of the GNN layer $l$. \\
        & $d^l_{in}$, $d^l_{out}$ &  The dimension of the input/output hidden vectors of the GNN layer $l$. \\
        & $dim(\MyVec{x})$ & The dimension of a vector $\MyVec{x}$. \\
        \bottomrule
    \end{tabular}
\end{table}

@@

A GNN usually consists of more than one GNN layers.
%
Each GNN layer consists of $|\mathcal{V}|$ graph neurons, where $|\mathcal{V}|$ is the number of vertices in $\mathcal{G}$.
%
Each graph neuron corresponds to a vertex in $\mathcal{G}$.
%
Different GNNs mainly differ in the graph neurons that they use.
%
We elaborate on details of graph neurons later.
%
GNN layers are \emph{sparsely} connected with the input layer and other GNN layers.
%
\begin{itemize}
    \item In the first GNN layer (layer 0), a graph neuron collects feature vectors from the input layer.
    %
    For the graph neuron corresponding to the vertex $v_x$, it collects the feature vector $\boldsymbol{v}_x$ and the feature vectors $\boldsymbol{v}_y$ of the vertices $v_y$ that are adjacent to $v_x$ (i.e.  $v_y \in \mathcal{N}(v_x)$).
    %
    The graph neuron aggregates input feature vectors, apply non-linear transformation, and outputs a hidden vector $\boldsymbol{h}^1_x$ for $v_x$.
    %
    Take the demo GNN in \figurename~\ref{fig:general_structure_of_gnn}(a) as an example.
    %
    Since $\mathcal{N}(v_3) = \{v_1, v_2, v_4, v_5, v_6\}$, the graph neuron of $v_3$ at the GNN layer 0 collects the input feature vectors \{$\boldsymbol{v}_1$, $\boldsymbol{v}_2$, $\boldsymbol{v}_3$, $\boldsymbol{v}_4$, $\boldsymbol{v}_5$, $\boldsymbol{v}_6$\} from the input layer and outputs $\boldsymbol{h}^1_3$.
    %
    The connections between the first GNN layer and the input layer  are determined by the topology of $\mathcal{G}$.
    %
    The graph neuron of $v_x$ in the first GNN layer is connected with the input unit of $v_y$  in the input layer only if there is an edge $e_{y,x}$ between $v_y$ and $v_x$ in $\mathcal{G}$.
    %
    Since most real-world graphs are very \emph{sparse} (i.e. $|\mathcal{E}| \ll |\mathcal{V}|^2$), the connections between the first GNN layer and the input layer are also sparse, different from the traditional neural networks.

@@

Supporting end-to-end training is a prominent advantage of GNNs, compared with traditional graph-based machine learning methods.
The traditional methods need to construct input feature vectors for vertices and edges manually or use embedding methods like DeepWalk \cite{bryan2014_deepwalk} and node2vec \cite{aditya2016_node2vec}.
The feature vector generation is independent from the model training.
Therefore, generated feature vectors may not be suitable for downstream prediction tasks.
In GNNs, gradients are propagated from the prediction layer back to GNN layers layer by layer. 
The model parameters in the GNN layers are updated based on the feedbacks from the downstream prediction task. 
In a fully parameterized way, a GNN can automatically extract an embedding vector for each vertex from its $L$-hop neighborhood, tuned according to the specific prediction task.

@@

The graph neuron of $v_x$ calculates the output hidden vector $\boldsymbol{h}^{l+1}_x$ with the three functions in three steps:
%
\begin{enumerate}
    %
    \item  For every adjacent edge $(v_y, v_x)$ of $v_x$ ($v_y \in \mathcal{N}(v_x)$), the messaging function $\phi^l$ receives the output hidden vectors $\boldsymbol{h}^l_x$ and $\boldsymbol{h}^l_y$ from the previous GNN layer and the edge feature vector $\boldsymbol{e}_{y,x}$ (and other feature vectors associated with $v_x$/$v_y$ if necessary) as the input.
    %
    $\phi^l$ emits a \emph{message vector} $\boldsymbol{m}^l_{y,x}$ for every edge $(v_y, v_x)$ at the layer $l$, i.e. $\boldsymbol{m}^l_{y,x} = \phi^l(\boldsymbol{h}^l_y, \boldsymbol{h}^l_x, \boldsymbol{e}_{y,x}, \dots)$;
    %
    \item The aggregation function $\Sigma^l$ then aggregates the message vectors $\boldsymbol{m}^l_{y,x}$ of the adjacent edges ($v_y \in \mathcal{N}(v_x)$) to produce an \emph{aggregated vector} $\boldsymbol{s}^l_x$, i.e. $\boldsymbol{s}^l_{x} = \sum^l_{v_y \in \mathcal{N}(v_x)}{\boldsymbol{m}^l_{y,x}}$;
    %
    \item The updating function $\gamma^l$ calculates the {hidden vector} of this layer $\boldsymbol{h}^{l+1}_x$ based on the hidden vector from the previous layer $\MyVec{h}^l_x$ and the aggregated vector $\MyVec{s}^l_x$ (and other feature vectors associated with $v_x$ if necessary), i.e. $\boldsymbol{h}^{l+1}_x = \gamma^l(\boldsymbol{h}^l_x, \boldsymbol{s}^l_x, \dots)$.
\end{enumerate}

@@

GGNN \cite{li2015_ggnn} introduces the gated recurrent unit (GRU) into graph neural networks.
%
The updating function $\phi^l$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
%
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
%
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
%
Since the messaging function $\phi^l$ only uses the hidden vector $\boldsymbol{h}^l_y$ of the source vertex $v_y$ for an edge $e_{y,x}$, in the implementation of PyG, GGNN conducts the pre-processing vertex calculation $\MyVec{\hat{h}}^l_x = \Param {\MyMat{W}^l}\boldsymbol{h}^l_x$ for every vertex $v_x$ before the message passing.
%
The messaging function $\phi^l$ directly uses $\MyVec{\hat{h}}^l_y$ as the message vector for every edge $e_{y, x}$.
%
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

@@

The neighbor sampling techniques \cite{hamilton2017_graphsage, ying2018_pinsage, chen2018_fastgcn, chen2018_sgcn, huang2018_adap} sample nodes or edges layer by layer.
%
The sampled subgraphs of different GNN layers may be \emph{different}, as shown in \figurename~\ref{fig:gnn_sampling}(a).
%
GraphSAGE \cite{hamilton2017_graphsage} is a representative neighbor sampling technique.
%
The tecnnique first samples several vertices from $\mathcal{V}$ in the last GNN layer.
%
Then it repeatedly samples the neighbors of the sampled vertices from the previous layer until the input layer.
%
For every sampled vertex $v_x$ in the GNN layer $l$, GraphSAGE samples at most $\eta^l$ neighbors of $v_x$ from the previous layer.
%
$\eta^l$ is the hyper-parameter that is usually much smaller than $|\mathcal{V}|$.
%
By this way, GraphSAGE limits neighborhood sizes of vertices in the sampled subgraph, especially high-degree vertices.

@@

The graph sampling techniques \cite{zeng2018_aesg, chiang2019_cluster_gcn, zeng2020_graphsaint} sample a subgraph for each mini-batch and use the same sampled subgraph for all GNN layers, as shown in \figurename~\ref{fig:gnn_sampling}(b).
%
They differ in how to sample subgraphs.
%
The cluster sampler technique \cite{chiang2019_cluster_gcn} is a representative graph sampling technique.
%
Given a training graph $\mathcal{G}$, it partitions $\mathcal{G}$ into several dense clusters.
%
For each mini-batch, it randomly pick $N$ clusters to form the sampled subgraph, where $N$ is the hyper-parameter.

@@

The model parameters of a GNN consist of the weight vectors/matrices in the prediction layer and the messaging/aggregation/updating functions of each GNN layer.
%
For transductive learning, the input graphs of training and inference are same.
%
The structure of the GNN does not need to adjust. 
%
For inductive learning, the input graphs used in training and inference are different.
%
When we want to perform inference on a new graph with a pre-trained GNN, the structure of the GNN needs to be adjusted, but the hyper-parameters and the model parameters of the GNN remain unchanged.
%
The number of graph neurons in each GNN layer has to be adjusted to $|\mathcal{V}|$ of the new graph.
%
The connections of graph neurons between GNN layers are also adjusted according to the edge set of the new graph.
%
During inference, input feature vectors of the input graph are propagated forward from the input layer to the prediction layer to give out predicted labels.

@@

The implementation of sample-based inference in PyG stores the whole input graph and its feature vectors in the main memory.
%
In the node classification task, it splits the vertices in the test set into batches accoding to a given batch size.
%
The subgraph corresponding to each batch is sampled on the CPU side and sent to the GPU side to perform inference.
%

>>>>

\begin{table}[h]
    \caption{Frequently-used symbols}
    \label{tab:notations}
    \centering
    \small
    \begin{tabular}{p{3em}lp{35em}}
        \toprule
       Category & Symbol & Meaning \\
        \midrule
        \multirow[c]{4}{3em}{Graph Structure}& $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ & The simple undirected input graph with the vertex set $\mathcal{V}$ and the edge set $\mathcal{E}$. \\
        & $v_x$ & The $x$-th vertex of the input graph. \\
        & $e_{y,x}$ & The edge pointing from $v_y$ to $v_x$ of the input graph. \\
        & $\mathcal{N}(v_x)$ & The adjacency set of $v_x$ in the input graph. \\ 
        & $\bar{d}$ & The average degree of the input graph. \\ \midrule
        \multirow[c]{6}{3em}{GNN Definition}& $L$ & The number of GNN layers. \\
        & $K$ & The number of heads in a GNN layer. \\
        & $\phi^l$ & The messaging function of the GNN layer $l$. \\
        & $\Sigma^l$ & The aggregation function of the GNN layer $l$. \\
        & $\gamma^l$ & The vertex updating function of the GNN layer $l$. \\ 
        & $\phi^{l,i}$ / $\Sigma^{l,i}$ / $\gamma^{l,i}$ & The messaging/aggregation/updating function of the $i$-th sub-layer of the GNN layer  $l$.\\
        & $\textcolor{blue}{\boldsymbol{W}^l, \boldsymbol{W}^{(k)}/\boldsymbol{b}, \boldsymbol{a}}$ & The matrices/vectors represented by the blue characters are the weight matrices/vectors that need to be learned in the GNN. \\  \midrule
        \multirow[c]{8}{3em}{Vector}& $\boldsymbol{v}_x$ & The feature vector of the vertex $v_x$. \\
        & $\boldsymbol{e}_{y,x}$ & The feature vector of the edge $e_{y,x}$.  \\
        & $\boldsymbol{h}_x^{l}$ &  The {input} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$. \\
        & $\boldsymbol{h}_x^{l+1}$ &  The {output} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$.\\
        & $\boldsymbol{m}_{y,x}^l$ & The message vector of the edge $e_{y,x}$ outputted by $\phi^l$ of the GNN layer $l$. \\
        & $\boldsymbol{s}_{x}^l$ & The aggregated vector of the vertex $v_x$ outputted by $\Sigma^l$ of the GNN layer $l$. \\
        & $\boldsymbol{h}_{x}^{l,i}$ / $\boldsymbol{m}_{y,x}^{l,i}$ / $\boldsymbol{s}_{x}^{l,i}$ & The hidden/message/aggregated vector of the vertex $v_x$ outputted by $\gamma^{l,i}$/$\phi^{l,i}$/$\Sigma^{l,i}$ of the $i$-th sub-layer of the GNN layer $l$. \\
        & $d^l_{in}$, $d^l_{out}$ &  The dimension of the input/output hidden vectors of the GNN layer $l$. \\
        & $dim(\MyVec{x})$ & The dimension of a vector $\MyVec{x}$. \\
        \bottomrule
    \end{tabular}
\end{table}

@@

A GNN usually consists of more than one GNN layer.
%
Each GNN layer consists of $|\mathcal{V}|$ graph neurons, where $|\mathcal{V}|$ is the number of vertices in $\mathcal{G}$.
%
Each graph neuron corresponds to a vertex in $\mathcal{G}$.
%
Different GNNs mainly differ in the graph neurons that they use.
%
We elaborate on the details of graph neurons later.
%
GNN layers are \emph{sparsely} connected with the input layer and other GNN layers.
%
\begin{itemize}
    \item In the first GNN layer (layer 0), a graph neuron collects feature vectors from the input layer.
    %
    For the graph neuron corresponding to the vertex $v_x$, it collects the feature vector $\boldsymbol{v}_x$ and the feature vectors $\boldsymbol{v}_y$ of the vertices $v_y$ that are adjacent to $v_x$ (i.e.  $v_y \in \mathcal{N}(v_x)$).
    %
    The graph neuron aggregates input feature vectors, apply non-linear transformation, and outputs a hidden vector $\boldsymbol{h}^1_x$ for $v_x$.
    %
    Take the demo GNN in \figurename~\ref{fig:general_structure_of_gnn}(a) as an example.
    %
    Since $\mathcal{N}(v_3) = \{v_1, v_2, v_4, v_5, v_6\}$, the graph neuron of $v_3$ at the GNN layer 0 collects the input feature vectors \{$\boldsymbol{v}_1$, $\boldsymbol{v}_2$, $\boldsymbol{v}_3$, $\boldsymbol{v}_4$, $\boldsymbol{v}_5$, $\boldsymbol{v}_6$\} from the input layer and outputs $\boldsymbol{h}^1_3$.
    %
    The connections between the first GNN layer and the input layer  are determined by the topology of $\mathcal{G}$.
    %
    The graph neuron of $v_x$ in the first GNN layer is connected with the input unit of $v_y$  in the input layer only if there is an edge $e_{y,x}$ between $v_y$ and $v_x$ in $\mathcal{G}$.
    %
    Since most real-world graphs are very \emph{sparse} (i.e. $|\mathcal{E}| \ll |\mathcal{V}|^2$), the connections between the first GNN layer and the input layer are also sparse, different from the traditional neural networks.

@@

Supporting end-to-end training is a prominent advantage of GNNs, compared with traditional graph-based machine learning methods.
The traditional methods need to construct input feature vectors for vertices and edges manually or use embedding methods like DeepWalk \cite{bryan2014_deepwalk} and node2vec \cite{aditya2016_node2vec}.
The feature vector generation is independent of model training.
Therefore, generated feature vectors may not be suitable for downstream prediction tasks.
In GNNs, gradients are propagated from the prediction layer back to GNN layers layer by layer. 
The model parameters in the GNN layers are updated based on the feedback from the downstream prediction task. 
In a fully parameterized way, a GNN can automatically extract an embedding vector for each vertex from its $L$-hop neighborhood, tuned according to the specific prediction task.

@@

The graph neuron of $v_x$ calculates the output hidden vector $\boldsymbol{h}^{l+1}_x$ with the three functions in three steps:
%
\begin{enumerate}
    %
    \item  For every adjacent edge $(v_y, v_x)$ of $v_x$ ($v_y \in \mathcal{N}(v_x)$), the messaging function $\phi^l$ receives the output hidden vectors $\boldsymbol{h}^l_y$ and $\boldsymbol{h}^l_x$ from the previous GNN layer and the edge feature vector $\boldsymbol{e}_{y,x}$ (and other feature vectors associated with $v_x$/$v_y$ if necessary) as the input.
    %
    $\phi^l$ emits a \emph{message vector} $\boldsymbol{m}^l_{y,x}$ for every edge $(v_y, v_x)$ at the layer $l$, i.e. $\boldsymbol{m}^l_{y,x} = \phi^l(\boldsymbol{h}^l_y, \boldsymbol{h}^l_x, \boldsymbol{e}_{y,x}, \dots)$;
    %
    \item The aggregation function $\Sigma^l$ then aggregates the message vectors $\boldsymbol{m}^l_{y,x}$ of the adjacent edges ($v_y \in \mathcal{N}(v_x)$) to produce an \emph{aggregated vector} $\boldsymbol{s}^l_x$, i.e. $\boldsymbol{s}^l_{x} = \sum^l_{v_y \in \mathcal{N}(v_x)}{\boldsymbol{m}^l_{y,x}}$;
    %
    \item The updating function $\gamma^l$ calculates the {hidden vector} of this layer $\boldsymbol{h}^{l+1}_x$ based on the hidden vector from the previous layer $\MyVec{h}^l_x$ and the aggregated vector $\MyVec{s}^l_x$ (and other feature vectors associated with $v_x$ if necessary), i.e. $\boldsymbol{h}^{l+1}_x = \gamma^l(\boldsymbol{h}^l_x, \boldsymbol{s}^l_x, \dots)$.
\end{enumerate}

@@

GGNN \cite{li2015_ggnn} introduces the gated recurrent unit (GRU) into graph neural networks.
%
The updating function $\gamma^l$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
%
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
%
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
%
Since the messaging function $\phi^l$ only uses the hidden vector $\boldsymbol{h}^l_y$ of the source vertex $v_y$ for an edge $e_{y,x}$, in the implementation of PyG, GGNN conducts the pre-processing vertex calculation $\MyVec{\hat{h}}^l_x = \Param {\MyMat{W}^l}\boldsymbol{h}^l_x$ for every vertex $v_x$ before the message passing.
%
The messaging function $\phi^l$ directly uses $\MyVec{\hat{h}}^l_y$ as the message vector for every edge $e_{y, x}$.
%
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

@@

The neighbor sampling techniques \cite{hamilton2017_graphsage, ying2018_pinsage, chen2018_fastgcn, chen2018_sgcn, huang2018_adap} sample nodes or edges layer by layer.
%
The sampled subgraphs of different GNN layers may be \emph{different}, as shown in \figurename~\ref{fig:gnn_sampling}(a).
%
GraphSAGE \cite{hamilton2017_graphsage} is a representative neighbor sampling technique.
%
The technique first samples several vertices from $\mathcal{V}$ in the last GNN layer.
%
Then it repeatedly samples the neighbors of the sampled vertices from the previous layer until the input layer.
%
For every sampled vertex $v_x$ in the GNN layer $l$, GraphSAGE samples at most $\eta^l$ neighbors of $v_x$ from the previous layer.
%
$\eta^l$ is the hyper-parameter that is usually small.
%
In this way, GraphSAGE limits neighborhood sizes of vertices in the sampled subgraph, especially high-degree vertices.

@@

The graph sampling techniques \cite{zeng2018_aesg, chiang2019_cluster_gcn, zeng2020_graphsaint} sample a subgraph for each mini-batch and use the same sampled subgraph for all GNN layers, as shown in \figurename~\ref{fig:gnn_sampling}(b).
%
They differ in how to sample subgraphs.
%
The cluster sampler technique \cite{chiang2019_cluster_gcn} is a representative graph sampling technique.
%
Given a training graph $\mathcal{G}$, it partitions $\mathcal{G}$ into several dense clusters.
%
For each mini-batch, it randomly picks $N$ clusters to form the sampled subgraph, where $N$ is the hyper-parameter.

@@

The model parameters of a GNN consist of the weight vectors/matrices in the prediction layer and the messaging/aggregation/updating functions of each GNN layer.
%
For transductive learning, the input graphs of training and inference are the same.
%
The structure of the GNN does not need to adjust. 
%
For inductive learning, the input graphs used in training and inference are different.
%
When we want to perform inference on a new graph with a pre-trained GNN, the structure of the GNN needs to be adjusted, but the hyper-parameters and the model parameters of the GNN remain unchanged.
%
The number of graph neurons in each GNN layer has to be adjusted to $|\mathcal{V}|$ of the new graph.
%
The connections of graph neurons between GNN layers are also adjusted according to the edge set of the new graph.
%
During inference, input feature vectors of the input graph are propagated forward from the input layer to the prediction layer to give out predicted labels.

@@

The implementation of sample-based inference in PyG stores the whole input graph and its feature vectors in the main memory.
%
In the node classification task, it splits the vertices in the test set into batches according to the given batch size.
%
The subgraph corresponding to each batch is sampled on the CPU side and sent to the GPU side to perform inference.
%
